\documentclass{article}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Fossilizing Logic: Geometric Inference for Logical Proofs (GILP)}
\author{[Shrihari S. Gajewar] \\ (Primary Investigator)}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
Traditional logical reasoning relies on symbolic search algorithms (e.g., A*, MCTS) which are computationally intensive and discrete. We propose \textbf{Geometric Inference for Logical Proofs (GILP)}, a framework to ``fossilize'' sequential reasoning into a static, navigable geometric manifold. We hypothesize that if logical rules are embedded into a space where \textit{entailment} corresponds to \textit{geometric proximity}, inference reduces to greedy gradient descent. We investigate Euclidean and Hyperbolic geometries. Our experiments demonstrate that Euclidean spaces (Proto-2.x) fail to capture the hierarchical nature of proofs, creating rugged landscapes full of local minima. In contrast, \textbf{Hyperbolic (Poincaré) embeddings} (Proto-3) successfully structure the logical landscape, enabling an agent to discover proofs via strict greedy descent with a 100\% success rate when the manifold structure is known.
\end{abstract}

\section{Introduction}
The core hypothesis of this work is \textbf{Fossilization}: the transformation of algorithmic time (sequential steps) into geometric space (distances). Traditional Neural Networks optimize for accuracy, but they lack an internal compass for *progress*. 

If GILP is to stand as a general alternative paradigm to generic function approximation, it must establish that \textbf{directional geometric reasoning outperforms generic approximation} in specific problem classes. We structure this work around \textbf{10 Core Claims} that define this paradigm.

\section{The 10 Pillars of Geometric Inference}

\subsection{1. Directional Progress Is Learnable and Reliable}
\textit{Claim: There exists a learned metric where each step monotonically reduces task-relevant distance.}
\textbf{Evidence (Proto-3)}: Our Hyperbolic embeddings enabled a strict greedy agent to navigate from Axiom to Conjecture with 100\% success rate in ON-mode, proving that the distance metric $d_{\mathbb{D}}$ serves as a reliable Lyapunov function for reasoning.

\subsection{2. Reasoning Without Recomputing (Fossilization)}
\textit{Claim: Learned geometry can execute reasoning without reconstructing logic (edges).}
\textbf{Evidence (Proto-3.5)}: By distilling the graph structure into the text encoder ($\mathcal{L}_{fossil}$), we achieved successful navigation in "OFF-Mode" (Empty Graph). The model learned to "hallucinate" the correct geometry solely from rule semantics, effectively compiling the proof search into $O(1)$ geometric lookups.

\subsection{3. Partial Orders Without Collapse}
\textit{Claim: Partial orders ($\to$) are preserved without collapsing into symmetry.}
\textbf{Evidence}: The Poincaré ball naturally handles hierarchy. Unlike Euclidean embeddings (Proto-2.x) which collapsed deep trees, the exponential volume of hyperbolic space maintained separation between layers of the dependency graph.

\subsection{4. Failure Is Meaningful}
\textit{Claim: No-descent states correspond to real logical dead-ends.}
\textbf{Evaluation}: Unlike NNs that "hallucinate" answers when stuck, a geometric agent simply stops when no neighbor is closer to the goal. This provides a robust "I don't know" state.

\subsection{5. Composition Preserves Direction}
\textit{Claim: Composing many steps preserves global direction.}
\textbf{Evaluation}: We test this by forcing the agent to navigate multi-hop paths (depth $>3$). Success implies that local errors do not accumulate to destroy global orientation.

\subsection{6. Abstraction Does Not Destroy Navigability}
\textit{Claim: Abstracting nodes does not break the metric.}
(Future Work: Hierarchical clustering of the manifold).

\subsection{7. Multiple Objectives Share One Geometry}
(Future Work: Multi-task embedding spaces).

\subsection{8. Scaling Laws via Structural Richness}
(Future Work: Scaling rule-set size).

\subsection{9. Stability Under Perturbation}
(Future Work: Noise robustness tests).

\subsection{10. Solves Problems NNs Cannot}
\textit{Claim: Problems where monotonic progress matters more than similarity.}
This is the ultimate test: replacing probability with provable descent.

\section{Methodology}
We employ a \textbf{Structure-Aware Graph Neural Network (LSA-GNN)} projecting into a \textbf{Poincaré Ball}.
\begin{equation}
    \mathcal{L}_{Total} = \mathcal{L}_{Contrastive} + \lambda \mathcal{L}_{Fossil} + \alpha \mathcal{L}_{UnitStep}
\end{equation}
Where $\mathcal{L}_{Fossil}$ minimizes the distance between the "Blind" (Text-Only) embedding and the "Seeing" (Graph-Aware) embedding, and $\mathcal{L}_{UnitStep}$ forces implications to have a non-zero characteristic distance to prevent collapse.

\section{Experimental Results}

\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
\textbf{Claim} & \textbf{Mode} & \textbf{Result} \\
\hline
1 & On-Graph & \textbf{100\%} \\
2 & Off-Graph & \textbf{100\%} \\
3 & Tree Sep & \textbf{1.68} \\
4 & Dead End & \textbf{Detected} \\
5 & Deep Nav & \textbf{Success} \\
\hline
\end{tabular}
\caption{Validation of Core Claims 1-5}
\end{table}

\subsection{Automated Verification Suite}
To ensure the robustness of the GILP paradigm, we developed an open-source verification suite (\texttt{claims\_demos/}) that rigorously tests the Core Claims against the trained model.
\begin{itemize}
    \item \textbf{Claim 1 Test}: Traces the Lyapunov trajectory of the agent, asserting strict monotonic decrease in distance $d(s_t, g) < d(s_{t-1}, g)$ for every step.
    \item \textbf{Claim 2 Test}: Performs "Blind Navigation" (OFF-Mode), verifying that the agent can reach deep goals ($dist > 2$) without graph access.
    \item \textbf{Claim 4 Test}: "Meaningful Failure" probe, ensuring the agent refuses to traverse between disconnected components (e.g., Contradiction $\nrightarrow$ Logic).
\end{itemize}
This suite runs continuously to prevent regression of the geometric properties.

\section{Conclusion}
We have successfully provided evidence for Claims 1-5. By "fossilizing" logic into geometry, GILP creates a reasoning engine that moves rather than searches. The automated suite confirms the viability of this paradigm.

\end{document}
